---
title: '**VarSelLCM**'
output:
  html_document: default
  pdf_document: default
---


*Variable Selection for Model-Based Clustering of Continuous, Count, Categorical or Mixed-Type Data Set with Missing Values.*

**Description:**
  
* *Authors*: **Matthieu Marbac** and **Mohammed Sedki**.
* *License*: [GPL-2](http://www.gnu.org/licenses/gpl-2.0.html).
* *Download VarSelLCM 2.0.1 (beta version for mixed-type data)*: [link](http://cran.r-project.org/web/packages/VarSelLCM/).  

**References:** 

* Marbac, M. and Sedki, M. (2017), [Variable selection for model-based clustering using the integrated complete-data likelihood](https://link.springer.com/article/10.1007/s11222-016-9670-1), Statistics and Computing, Volume 27, Issue 4, pp 1049–1063.
* Marbac, M., Patin, E. and Sedki, M. (2018), [Variable selection for mixed data clustering: Application in human population genomics](https://arxiv.org/abs/1703.02293), Arxiv 1703.02293.

**Introduction:**  
*VarSelLCM* permits model selection (number of component and variable selection) for clustering.

It can analyze continuous, categorical, integer and mixed-data. Moreover, data with missing values can be analyze without any pre-processing.

A Shiny application is implemented to easily interpret the clustering results.

Imputation function is implemented. Thus, *VarSelLCM* can be used as an imputation approach based on mixture model.

Tool functions (*summary, print* and *plot*) facilitate the result interpretation.




**Overview of the VarSelLCM functions**

This section performs the whole analysis of the *Heart* data set . It uses all the functions implemented in the package *VarSelLCM* and can be used as a tutorial. 

Clustering is performed with two clusters by doing a variable selection. Model selection is done with BIC.

**Loadings**

```{r, comment=""}
library(VarSelLCM)
```


```{r, comment=""}
# Data loading
data("heart")
# Add a missing value (just to show that it works!)
heart[1,1] <- NA
# Clustering with variable selection and a number of cluster betwee 1 and 4
# Model selection is BIC (to use MICL, the option must be specified)
out <- VarSelCluster(heart[,-13], 1:4)
# Summary of the best model
summary(out)
# Discriminative power of the variables (here, the most discriminative variable is MaxHeartRate)
plot(out, type="bar")
# Boxplot for continuous (or interger) variable
plot(out, y="MaxHeartRate", type="boxplot")
# Empirical and theoretical distributions (to check that clustering is pertinent)
plot(out, y="MaxHeartRate", type="cdf")
# Summary of categorical variable
plot(out, y="Sex")
# Summary of the probabilities of missclassification
plot(out, type="probs-class")
```

```{r, eval=FALSE, comment="", include=TRUE}
# Start the shiny application
 VarSelShiny(out)
```